```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
There are 3-4 packages you will need to install for today's practical: `install.packages(c("xgboost", "eegkit", "forecast", "tseries", "caret"))` apart from that everything else should already be available on your system. 

If you are using a newer Mac you may have to also install [quartz](https://www.xquartz.org/) to have everything work (do this if you see errors about `X11` during install/execution).

I will endeavour to use explicit imports to make it clear where functions are coming from (functions without `library_name::` are part of base R or a function we've defined in this notebook).

```{r libraries, echo=FALSE}
# Using the same library we used earlier in the course for tabular data because we know it works!
library(xgboost)

# EEG manipulation library in R (although very limited compared to signal processing libraries available in other languages, matlab might actually still be a leader in this specific area)
library(eegkit)

# some time series functions (that we only skim the depths of)
library(forecast)
library(tseries)
library(caret)

# just tidyverse libraries that should already be installed
library(dplyr)
library(reshape2)
library(purrr)
library(ggplot2)
```

## EEG Eye Detection Data

One of the most common types of medical sensor data (and one that we talked about during the lecture) are Electroencephalograms (EEGs).  
These measure mesoscale electrical signals (measured in microvolts) within the brain, which are indicative of a region of neuronal activity.
Typically, EEGs involve an array of sensors (aka channels) placed on the scalp with a high degree of covariance between sensors.

As EEG data can be very large and unwieldy, we are going to use a relatively small/simple dataset today from [this paper](http://ehrai.com/su/pdf/aihls2013.pdf).

This dataset is a 117 second continuous EEG measurement collected from a single person with a device called a "Emotiv EEG Neuroheadset".
In combination with the EEG data collection, a camera was used to record whether person being recorded had their eyes open or closed. 
This was eye status was then manually annotated onto the EEG data with `1` indicated the eyes being closed and `0` the eyes being open.
Measures microvoltages are listed in chronological order with the first measured value at the top of the dataframe.

Let's parse the data directly from the `h2o` library's (which we aren't actually using directly) test data S3 bucket:

```{r parse_data}
eeg_url <- "https://h2o-public-test-data.s3.amazonaws.com/smalldata/eeg/eeg_eyestate_splits.csv"
eeg_data <- read.csv(eeg_url)

# add timestamp
Fs <- 117 / nrow(eeg_data)
eeg_data <- transform(eeg_data, ds = seq(0, 116.99999, by = Fs), eyeDetection = as.factor(eyeDetection))
print(table(eeg_data$eyeDetection))

# split dataset into train, validate, test
eeg_train <- subset(eeg_data, split == 'train', select = -split)
print(table(eeg_train$eyeDetection))

eeg_validate <- subset(eeg_data, split == 'valid', select = -split)
eeg_test <- subset(eeg_data, split == 'test', select = -split)
```

**0** Knowing the `eeg_data` contains 117 seconds of data, inspect the `eeg_data` dataframe and the code above to and determine how many samples per second were taken?

```{r}
SPS <- 117 / nrow(eeg_data)
```

There were 0.0078 samples per second taken

**1** How many EEG electrodes/sensors were used?

```{r}
num_electrodes <- ncol(eeg_data) - 2
```

There were 15 EEG electrodes/samples used

### Exploratory Data Analysis

Now that we have the dataset and some basic parameters let's begin with the ever important/relevant exploratory data analysis.

First we should check there is no missing data!
```{r check_na}
sum(is.na(eeg_data))
```

Great, now we can start generating some plots to look at this data within the time-domain.

First we use `reshape2::melt()` to transform the `eeg_data` dataset from a wide format to a long format expected by `ggplot2`.

Specifically, this converts from "wide" where each electrode has its own column, to a "long" format, where each observation has its own row. 
This format is often more convenient for data analysis and visualization, especially when dealing with repeated measurements or time-series data.

We then use `ggplot2` to create a line plot of electrode intensities per sampling time, with the lines coloured by electrode, and the eye status annotated using dark grey blocks.

```{r plot_data}
melt <- reshape2::melt(eeg_data %>% dplyr::select(-split), id.vars=c("eyeDetection", "ds"), variable.name = "Electrode", value.name = "microvolts")


ggplot2::ggplot(melt, ggplot2::aes(x=ds, y=microvolts, color=Electrode)) + 
  ggplot2::geom_line() + 
  ggplot2::ylim(3500,5000) + 
  ggplot2::geom_vline(ggplot2::aes(xintercept=ds), data=dplyr::filter(melt, eyeDetection==1), alpha=0.005)
```

**2** Do you see any obvious patterns between eyes being open (dark grey blocks in the plot) and the EEG intensities?

There seems to be a spike when the eye are open and a relatively relaxed state when the eyes are closed with the EEG intensities.

**3** Similarly, based on the distribution of eye open/close state over time to anticipate any temporal correlation between these states?

Based on the distribution of eye open/close state over time there is a temporal correlation between the states. There is more activity when the eyes are open as opposed to when they are closed.

Let's see if we can directly look at the distribution of EEG intensities and see how they related to eye status.


As there are a few extreme outliers in voltage we will use the `dplyr::filter` function to remove values outwith of 3750 to 50003. The function uses the `%in%` operator to check if each value of microvolts is within that range. The function also uses the `dplyr::mutate()` to change the type of the variable eyeDetection from numeric to a factor (R's categorical variable type).

```{r compare_distrib}
melt_train <- reshape2::melt(eeg_train, id.vars=c("eyeDetection", "ds"), variable.name = "Electrode", value.name = "microvolts")

# filter huge outliers in voltage
filt_melt_train <- dplyr::filter(melt_train, microvolts %in% (3750:5000)) %>% dplyr::mutate(eyeDetection=as.factor(eyeDetection))

ggplot2::ggplot(filt_melt_train, ggplot2::aes(y=Electrode, x=microvolts, fill=eyeDetection)) + ggplot2::geom_boxplot()
```



Plots are great but sometimes so it is also useful to directly look at the summary statistics and how they related to eye status.
We will do this by grouping the data based on eye status and electrode before calculating the statistics using the convenient `dplyr::summarise` function.

```{r compare_summary_stats}
filt_melt_train %>% dplyr::group_by(eyeDetection, Electrode) %>% 
    dplyr::summarise(mean = mean(microvolts), median=median(microvolts), sd=sd(microvolts)) %>% 
    dplyr::arrange(Electrode)
```




**4** Based on these analyses are any electrodes consistently more intense or varied when eyes are open?

AF3 and F7 seem to be the electrodes that experience are consistently more intense when the eyes are open.

#### Time-Related Trends

As it looks like there may be a temporal pattern in the data we should investigate how it changes over time.  

First we will do a statistical test for stationarity:

```{r convert_to_tseries}
apply(eeg_train, 2, tseries::adf.test)
```


**5** What is stationarity?

Stationarity is a property where characteristics remain stable over time.

**6** Why are we interested in stationarity? What do the results of these tests tell us? (ignoring the lack of multiple comparison correction...)

We are interested in stationarity because it makes modelling easier to do and ensures the parameters are accurate as well as makes the data more reliable.

The results of this test tell us that all the alternate hypotheses are stationary.

Then we may want to visually explore patterns of autocorrelation (previous values predict future ones) and cross-correlation (correlation across channels over time) using `forecast::ggAcf` function.

The ACF plot displays the cross- and auto-correlation values for different lags (i.e., time delayed versions of each electrode's voltage timeseries) in the dataset. 
It helps identify any significant correlations between channels and observations at different time points. 
Positive autocorrelation indicates that the increase in voltage observed in a given time-interval leads to a proportionate increase in the lagged time interval as well.
Negative autocorrelation indicates the opposite!


```{r correlation}
forecast::ggAcf(eeg_train %>% dplyr::select(-ds))
```





**7** Do any fields show signs of strong autocorrelation (diagonal plots)? Do any pairs of fields show signs of cross-correlation? Provide examples.

Some examples are (F7, F7), (F7, FC5), (Detect, detect), (FC6, FC6) 


#### Frequency-Space 

We can also explore the data in frequency space by using a Fast Fourier Transform.  
After the FFT we can summarise the distributions of frequencies by their density across the power spectrum.
This will let us see if there any obvious patterns related to eye status in the overall frequency distributions.

```{r fft_open}
eegkit::eegpsd(eeg_train %>% dplyr::filter(eyeDetection == 0) %>% dplyr::select(-eyeDetection, -ds), Fs = Fs, xlab="Eye Open")
```

```{r fft_closed}
eegkit::eegpsd(eeg_train %>% dplyr::filter(eyeDetection == 1) %>% dplyr::select(-eyeDetection, -ds), Fs = Fs, xlab="Eye Closed")
```




**8** Do you see any differences between the power spectral densities for the two eye states? If so, describe them.

Yes I notice some differences between the power spectral densities for the two eye states.

we notice firstly there is one more power spectral density in the eyes closed state.

it seems in the eyes open the density trends upwards while it is downwards for the eyes closed state.

#### Independent Component Analysis

We may also wish to explore whether there are multiple sources of neuronal activity being picked up by the sensors.  
This can be achieved using a process known as independent component analysis (ICA) which decorrelates the channels and identifies the primary sources of signal within the decorrelated matrix.

```{r ica, warning=FALSE}
ica <- eegkit::eegica(eeg_train %>% dplyr::select(-eyeDetection, -ds), nc=3, method='fast', type='time')
mix <- dplyr::as_tibble(ica$M)
mix$eyeDetection <- eeg_train$eyeDetection
mix$ds <- eeg_train$ds

mix_melt <- reshape2::melt(mix, id.vars=c("eyeDetection", "ds"), variable.name = "Independent Component", value.name = "M")


ggplot2::ggplot(mix_melt, ggplot2::aes(x=ds, y=M, color=`Independent Component`)) + 
  ggplot2::geom_line() + 
  ggplot2::geom_vline(ggplot2::aes(xintercept=ds), data=dplyr::filter(mix_melt, eyeDetection==1), alpha=0.005) +
  ggplot2::scale_y_log10()
```



**9** Does this suggest eye opening relates to an independent component of activity across the electrodes?

Yes this suggests eye opening relates to an independent component of activity across the electrodes. 

### Eye Opening Prediction

Now that we've explored the data let's use a simple model to see how well we can predict eye status from the EEGs:

```{r xgboost}
# Convert the training and validation datasets to matrices
eeg_train_matrix <- as.matrix(dplyr::select(eeg_train, -eyeDetection, -ds))
eeg_train_labels <- as.numeric(eeg_train$eyeDetection) -1

eeg_validate_matrix <- as.matrix(dplyr::select(eeg_validate, -eyeDetection, -ds))
eeg_validate_labels <- as.numeric(eeg_validate$eyeDetection) -1

# Build the xgboost model
model <- xgboost(data = eeg_train_matrix, 
                 label = eeg_train_labels,
                 nrounds = 100,
                 max_depth = 4,
                 eta = 0.1,
                 objective = "binary:logistic")

print(model)
```



**10** Using the `caret` library (or any other library/model type you want such as a naive Bayes) fit another model to predict eye opening.

```{r model2}

library(caret)
library(dplyr)

eeg_train_matrix <- as.matrix(dplyr::select(eeg_train, -eyeDetection, -ds))
eeg_train_labels <- as.factor(eeg_train$eyeDetection)

eeg_validate_matrix <- as.matrix(dplyr::select(eeg_validate, -eyeDetection, -ds))
eeg_validate_labels <- as.factor(eeg_validate$eyeDetection)

train_data <- data.frame(eeg_train_matrix)
train_data$eyeDetection <- eeg_train_labels

set.seed(123)  
logistic_model <- train(eyeDetection ~ ., data = train_data, method = "glm", family = "binomial")

print(logistic_model)

validate_data <- data.frame(eeg_validate_matrix)
validate_data$eyeDetection <- eeg_validate_labels

predictions <- predict(logistic_model, newdata = validate_data)

confusion_matrix <- confusionMatrix(predictions, validate_data$eyeDetection)
print(confusion_matrix)

```


**11** Using the best performing of the two models (on the validation dataset) calculate and report the test performance (filling in the code below):

```{r test}
# Load necessary libraries
library(xgboost)
library(caret)
library(dplyr)

eeg_train_matrix <- as.matrix(dplyr::select(eeg_train, -eyeDetection, -ds))
eeg_train_labels <- as.numeric(eeg_train$eyeDetection) - 1

eeg_validate_matrix <- as.matrix(dplyr::select(eeg_validate, -eyeDetection, -ds))
eeg_validate_labels <- as.factor(eeg_validate$eyeDetection)

eeg_test_matrix <- as.matrix(dplyr::select(eeg_test, -eyeDetection, -ds))
eeg_test_labels <- as.factor(eeg_test$eyeDetection)

model <- xgboost(data = eeg_train_matrix, 
                 label = eeg_train_labels,
                 nrounds = 100,
                 max_depth = 4,
                 eta = 0.1,
                 objective = "binary:logistic")

xgb_predictions <- predict(model, eeg_validate_matrix)
xgb_predictions_binary <- ifelse(xgb_predictions > 0.5, 1, 0)

xgb_confusion_matrix <- confusionMatrix(as.factor(xgb_predictions_binary), eeg_validate_labels)
xgb_accuracy <- xgb_confusion_matrix$overall['Accuracy']

train_data <- data.frame(eeg_train_matrix)
train_data$eyeDetection <- as.factor(eeg_train_labels)

set.seed(123) 
logistic_model <- train(eyeDetection ~ ., data = train_data, method = "glm", family = "binomial")

validate_data <- data.frame(eeg_validate_matrix)
validate_data$eyeDetection <- eeg_validate_labels

logistic_predictions <- predict(logistic_model, newdata = validate_data)

logistic_confusion_matrix <- confusionMatrix(logistic_predictions, validate_data$eyeDetection)
logistic_accuracy <- logistic_confusion_matrix$overall['Accuracy']

if (logistic_accuracy > xgb_accuracy) {
  best_model <- logistic_model
  best_model_name <- "Logistic Regression"
  test_predictions <- predict(best_model, newdata = data.frame(eeg_test_matrix))
  test_confusion_matrix <- confusionMatrix(test_predictions, eeg_test_labels)
} else {
  best_model <- model
  best_model_name <- "XGBoost"
  test_predictions <- predict(best_model, eeg_test_matrix)
  test_predictions_binary <- ifelse(test_predictions > 0.5, 1, 0)
  test_confusion_matrix <- confusionMatrix(as.factor(test_predictions_binary), eeg_test_labels)
}

cat("Best Model:", best_model_name, "\n")
print(test_confusion_matrix)

```

**12** Describe 2 possible alternative modeling approaches for prediction of eye opening from EEGs we discussed in the lecture but haven't explored in this notebook.

A machine learning approach could be used where the features are directly extracted and trained on EEG signals to predict the opening and closed states of the eyes.

We could also make use of neural networks to automatically detect and accurately predict the states of the eye directly from data gotten from the EEG data.

**13** What are 2 R libraries you could use to implement these approaches? (note: you don't actually have to implement them though!)

Two R libraries that could be used to implement these approaches are Recurrent Neural Networks and Random Forests.

## Optional

**14** (Optional) As this is the last practical of the course - let me know how you would change future offerings of this course. This will not impact your marks!

I think also some more background explanation of the practical material before diving into doing it as I believe at least for myself I have little to no medical/health data experience or knowledge so to do these lots of time spent was on trying to understand as well as searching up how to properly analyze the data.

- What worked and didn’t work for you (e.g., in terms of the practicals, tutorials, and lectures)?

The presentation portion is very interesting having so to speak the self directed studies to find a gap in the space to explore and do research on is very interesting.

The practicals although for the most part were interesting I think they could be explained abit better, I found as someone who had little to no experience in analyzing health data I found myself lost more often than not when dealing with the practicals.

The lectures were interesting, good topic analysis as well as very varied in the space. I think the slides could be trimmed down abit for easier digestion of the content.

- Was learning how to run the practicals on your own machines instead of a clean server that will disappear after the course worth the technical challenges?

It is good to have them on our machines as they could be looked back on later down the lines as the practicals were pretty interesting and the testing could be done locally without having to rely on external factors.
 
- What would you add or remove from the course? 

Nothing really I think it is well structured the readings were interesting and gave a fun perspective of things. The practicals were also fun although I will add I think they need some more clarifications in some cases.

- What was the main thing you will take away from this course?

Analyzing health data is interesting and there is alot more that goes into getting useful data from basic things.